---
title: "Data exploration and cleaning"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---

This notebook performs a basic overview of the data. It uses as input the "clean kobo output" that was first cleaned by `1_quality_check_and_cleaning.Rmd`.

## Get data

Load required libraries:

```{r, warning=FALSE, message=FALSE}
library(tidyr)
library(dplyr)
library(utile.tools)
library(ggplot2)
library(ggsankey)
```

Load required functions. These custom fuctions are available at: https://github.com/AliciaMstt/GeneticIndicators

```{r}
source("get_indicator1_data.R")
source("get_indicator2_data.R")
```


Get indicator 1 and indicator 2 data from clean kobo output
```{r}
# Get data for each indicator:
ind1_data<-get_indicator1_data(file="")
head(ind1_data)

ind2_data<-get_indicator2_data(file="")
head(ind2_data)

```

Filter out records which were marked as "not_approved" in the manual Kobo validation interface (this means country assessors determined the is something wrong with that particular record).

```{r}
# check if any species is flagged as "validation_status_not_approved"
ind2_data %>%
      filter(X_validation_status=="validation_status_not_approved")

# omit those records from data:
ind2_data<- ind2_data %>%
            filter(X_validation_status!="validation_status_not_approved")

ind1_data<- ind1_data %>%
            filter(X_validation_status!="validation_status_not_approved")

kobo_output<-kobo_output %>%
            filter(X_validation_status!="validation_status_not_approved")
```



## General description of the dataset

Records by country

```{r}
ggplot(kobo_output, aes(x=country_assessment)) + 
  geom_bar(stat = "count")

```

Records by taxonomic groups

```{r}
ggplot(kobo_output, aes(x=taxonomic_group)) + 
  geom_bar(stat = "count") +
  theme(axis.text.x = element_text(angle = 45))

```

Which taxonomic groups are countries assesing?

```{r}
library(ggsankey)

# transform data to how ggsankey wants it
df <- kobo_output %>%
  make_long(country_assessment, taxonomic_group)
df

# Sankey plot
ggplot(df, aes(x = x,
               next_x = next_x,
               node = node,
               next_node = next_node,
               fill = factor(node),
               label = node)) +
  geom_sankey(flow.alpha = 0.6, 
              node.color = NA,
              show.legend = FALSE) +
  geom_sankey_label(size = 2, color = 1, fill = "white") +
  # colour by country
  scale_fill_manual(values=c(scales::hue_pal()(length(unique(kobo_output$country_assessment))),  # a nice  color for each of the n countries
                             rep("darkgrey", length(unique(kobo_output$taxonomic_group)))), # grey for the n taxonomic groups
                    breaks=c(unique(kobo_output$country_assessment), 
                             unique(kobo_output$taxonomic_group))) +
  theme_sankey(base_size = 10) +
  xlab("")
  
```


How is the distribution of Ne / Nc data across countries?
```{r}
# transform data to how ggsankey wants it
df <- kobo_output %>%
  make_long(country_assessment, popsize_data)
df

# Sankey plot
ggplot(df, aes(x = x,
               next_x = next_x,
               node = node,
               next_node = next_node,
               fill = factor(node),
               label = node)) +
  geom_sankey(flow.alpha = 0.6, 
              node.color = NA,
              show.legend = FALSE) +
  geom_sankey_label(size = 2, color = 1, fill = "white") +
  # colour by country
  scale_fill_manual(values=c(scales::hue_pal()(length(unique(kobo_output$country_assessment))),  # a nice  color for each of the n countries
                             rep("darkgrey", length(unique(kobo_output$popsize_data)))), # grey for the n taxonomic groups
                    breaks=c(unique(kobo_output$country_assessment), 
                             unique(kobo_output$popsize_data))) +
  theme_sankey(base_size = 10) +
  xlab("")
  
```




How is Ne / Nc  data distributred across countries and speices?
```{r}
# transform data to how ggsankey wants it
df <- kobo_output %>%
  make_long(country_assessment, taxonomic_group, popsize_data, ne_pops_exists)
df

# Sankey
ggplot(df, aes(x = x,
               next_x = next_x,
               node = node,
               next_node = next_node,
               fill = factor(node),
               label = node)) +
  geom_sankey(flow.alpha = 0.6, 
              node.color = NA,
              show.legend = FALSE) +
  geom_sankey_label(size = 2, color = 1, fill = "white") +
  theme_sankey(base_size = 10) +
  xlab("")

```


