---
title: "Plots by individual country"
output: html_document
date: "2023-08-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Packages and functions

Load required libraries:

```{r, warning=FALSE, message=FALSE}
library(tidyr)
library(dplyr)
library(utile.tools)
library(stringr)
library(ggplot2)
library(ggsankey)
library(alluvial)
library(viridis)
library(cowplot)
```

Load required functions. These custom fuctions are available at: https://github.com/AliciaMstt/GeneticIndicators

```{r source}
source("get_indicator1_data.R")
source("get_indicator2_data.R")
source("get_indicator3_data.R")
source("get_metadata.R")
source("transform_to_Ne.R")
source("estimate_indicator1.R")
```

Other custom functions:
```{r custom_funs}
### not in
'%!in%' <- function(x,y)!('%in%'(x,y))


#' Duplicates data to create additional facet. Thanks to https://stackoverflow.com/questions/18933575/easily-add-an-all-facet-to-facet-wrap-in-ggplot2
#' @param df a dataframe
#' @param col the name of facet column
#'  
CreateAllFacet <- function(df, col){
  df$facet <- df[[col]]
  temp <- df
  temp$facet <- "all"
  merged <-rbind(temp, df)

  # ensure the facet value is a factor
  merged[[col]] <- as.factor(merged[[col]])

  return(merged)
}
```

Custom colors:
```{r custom colors}
## IUCN official colors
# Assuming order of levels is: "re", "cr", "en", "vu", "nt", "lc", "dd", "not_assessed", "unknown" (for regional, and w/o "re" for global). Make sure to change the levels to that order before plotting.
IUCNcolors<-c("brown2", "darkorange", "yellow", "green", "darkgreen", "darkgrey", "azure2", "bisque1")
IUCNcolors_regional<-c("darkorchid2", "brown2", "darkorange", "yellow", "green", "darkgreen", "darkgrey", "azure2", "bisque1")

## nice soft ramp for taxonomic groups
taxoncolors<-cividis(12) # same than using cividis(length(levels(as.factor(metadata$taxonomic_group))))
  
```

## Get data 
Get indicators and metadata data (single file with all), as well as indicator 1 population raw data (since this is by population and the metadata and indicators are estimated by species, indicator 1 raw data is in a different file):
```{r, echo=TRUE}
# Get data:
indicators_full<-read.csv(file="indicators_full.csv", header=TRUE)
ind1_data<-read.csv(file="ind1_data.csv", header=TRUE)

```

Select which country you want to analyse:
```{r}
# after "desired_country" in the code below, write the name of your country as it is written here:
# "australia"     "belgium"       "colombia"      "france"        "japan"         "mexico"        "south_africa"  "sweden"       
# "united_states"

desired_country<-"belgium"
```


```{r}
# Subset data to desired country:
indicators_full <- indicators_full %>% 
                   filter(country_assessment==desired_country)

ind1_data<-ind1_data %>% filter(country_assessment==desired_country)
```


## General description of the dataset

`r desired_country` has a total of `r nrow(indicators_full)` records. Of those, some taxa could have been assessed more than once, for example to account for different methods to define populations.

To explore what kind of taxa the country assessed regardless of if they assessed them once or more, lets create a dataset keeping all single assessed taxa, plus only the first assessment for taxa assessed multiple times. 

```{r}
# object with single assessed taxa, plus only the first assessment for taxa assessed multiple times
firstmulti<-indicators_full[!duplicated(cbind(indicators_full$taxon, indicators_full$country_assessment)), ]

```

How many taxa were assessed (i.e. counting only once taxa that were assessed multiple times)?
```{r}
# how many?
nrow(firstmulti)
```

### Taxonomic groups
Records by country, including taxa assessed more than once (see below for details on this)

```{r, out.width="700px", out.height="400px"}
ggplot(indicators_full, aes(x=taxonomic_group)) + 
  geom_bar(stat = "count") +
  xlab("") +
  ggtitle("Number of taxa by taxonomic groups assessed") +
  theme_light()

```

### Population size data availability

Population size (Nc or Ne) data availability at the taxon level:
```{r}
ggplot(indicators_full, aes(x=taxonomic_group, fill=popsize_data)) + 
  geom_bar(stat = "count") +
  coord_flip() +
  scale_fill_manual(values=c("#2ca02c", "#1f77b4", "grey80"),
                    breaks=c("yes", "data_for_species", "insuff_data_species"),
                    labels=c("Population level", "Species or subspecies level", "Insufficient data")) +
  labs(fill="Population size data availability",
       x="",
       y="Number of taxa (including records of taxa assessed more than once)") +
  theme_light() +
  theme(panel.border = element_blank(), legend.position="top")
  
```
Ne available?
```{r}
indicators_full %>% 
  filter(!is.na(ne_pops_exists)) %>% 
  filter(ne_pops_exists!="other_genetic_info") %>%
    ggplot(aes(x=taxonomic_group, fill=ne_pops_exists)) + 
  geom_bar() +
  coord_flip() +
scale_fill_manual(labels=c("no", "yes"),
                      breaks=c("no_genetic_data", "ne_available"),
                      values=c("#ff7f0e", "#2ca02c")) +
xlab("") +
ylab("Number of taxa") +
labs(fill="Ne available \n(from genetic data)")  +
theme_light() +
theme(text = element_text(size = 14), legend.position = "right", panel.border = element_blank())


```

Nc data available by taxa?
```{r}
indicators_full %>%
  filter(!is.na(nc_pops_exists)) %>%
    ggplot(aes(x=taxonomic_group, fill=nc_pops_exists)) +
    geom_bar() +
    coord_flip() +
    scale_fill_manual(values=c("#ff7f0e", "#2ca02c")) + 
    labs(fill="Nc available") +
    xlab("") +
    ylab("Number of taxa") +
    theme_light() +
    theme(text = element_text(size = 14), legend.position = "right", panel.border = element_blank())
```

What kind of Nc data?
```{r}
ind1_data %>%
  filter(!is.na(NcType)) %>%
  ggplot(aes(x=taxonomic_group, fill=NcType))+
  geom_bar() +
  scale_fill_manual(labels=c("Point", "Range \nor qualitative", "Unknown"),
                      breaks=c("Nc_point", "Nc_range", "unknown"),
                      values=c("#0072B2", "#E69F00", "grey80")) +
  xlab("") +
  ylab("Number of populations") +
  coord_flip() +
  labs(fill="Type of Nc data \nby population") +
  theme_light() +
  theme(text = element_text(size = 14), legend.position = "right", panel.border = element_blank())
 
```



## Methods used to define populations
Frequency table of methods used to define population
```{r}
table(indicators_full$defined_populations)
```

### Plot number of populations by method
```{r, out.height="500px", out.width="1064px"}
# Prepare data for plot with nice labels:
# sample size of TOTAL populations
sample_size <- indicators_full %>%
                    filter(!is.na(indicator2)) %>% 
                    group_by(defined_populations) %>% summarize(num=n())

# custom axis
## new dataframe
df<-indicators_full %>% 
  filter(n_extant_populations<500) %>%
    # add sampling size 
  left_join(sample_size) %>%
  mutate(myaxis = paste0(defined_populations, " (n= ", num, ")"))


# plot for number of pops
df %>%
  ggplot(aes(x=myaxis, y=n_extant_populations, color=defined_populations)) +
          geom_boxplot() + xlab("") + ylab("Number of mantained populations") +
          geom_jitter(size=.4, width = 0.1, color="black") +
  coord_flip() +
  theme_light() +
  theme(panel.border = element_blank(), legend.position="none",
        plot.margin = unit(c(0, 0, 0, 0), "cm")) + # this is used to decrease the space between plots
  scale_x_discrete(limits=rev) +  
  theme(text = element_text(size = 13))

```

### Plot indicator 2 by method
```{r}
# Prepare data for plot with nice labels:
# sample size of TOTAL populations
sample_size <- indicators_full %>%
                    filter(!is.na(indicator2)) %>% 
                    group_by(defined_populations) %>% summarize(num=n())

# custom axis
## new dataframe
df<-indicators_full %>% 
  filter(n_extant_populations<500) %>%
    # add sampling size 
  left_join(sample_size) %>%
  mutate(myaxis = paste0(defined_populations, " (n= ", num, ")"))



## plot for indicator 2
df %>%
  filter(n_extant_populations<500) %>%
  ggplot(aes(x=myaxis, y=indicator2, color=defined_populations)) +
          geom_boxplot() + xlab("") + ylab("Indicator 2") +
          geom_jitter(size=.4, width = 0.1, color="black") +
  coord_flip() +
  theme_light() +
  theme(panel.border = element_blank(), legend.position="none") +
  scale_x_discrete(limits=rev) +
  theme(text = element_text(size = 13))


```
### Plot Scatter plot of indicator2 and extant pops
```{r}
indicators_full %>%
  # filter outliers with too many pops
  # filter(n_extant_populations<200) %>%
  
  # plot
    ggplot(aes(x=n_extant_populations, y=indicator2, color=defined_populations)) +
    geom_point() +
    theme_light() +
    theme(legend.position = "none") +
    ylab("Indicator 2") +
    xlab("Number of mantained populations") +
    theme(text = element_text(size = 13))
```

## Missing data on extant and extinct populations

We have NA in indicator 2 because in some cases the number of extinct populations is unknown, therefore the operation cannot be computed. 

### Counts
Total records with NA in extant populations:
```{r}
sum(is.na(indicators_full$n_extant_populations))
```

Taxa with NA in extant populations:
```{r}
indicators_full %>%
  filter(is.na(n_extant_populations)) %>%
    select(taxonomic_group, taxon, n_extant_populations, n_extint_populations)
```

Total taxa with NA in **extinct** populations:
```{r}
sum(is.na(indicators_full$n_extint_populations))
```

Do taxa with NA for extant also have NA for extinct?

```{r}
indicators_full$taxon[is.na(indicators_full$n_extant_populations)] %in% indicators_full$taxon[is.na(indicators_full$n_extint_populations)]
```

### Plot missing data extinct populations
```{r}
indicators_full %>%
  ggplot(aes(x=taxonomic_group, fill=is.na(n_extint_populations))) +
  geom_bar() +
  scale_fill_manual(labels=c("number of populations known", "missing data"),
                    values=c("#2ca02c", "#ff7f0e")) + 
  coord_flip() +
  labs(fill="Extinct populations") +
  xlab("") + ylab("Number of taxa") +
  theme_light() +
  theme(panel.border = element_blank())
```


## Indicator 1 by type of range

Indicator 1 by type of range in the entire dataset:

```{r indicator1 by range type}
# get sample size by desired category
sample_size <- indicators_full  %>%
                    filter(!is.na(indicator1)) %>% 
                    group_by(species_range) %>% summarize(num=n())

# plot
indicators_full  %>% 
  # add sampling size 
  left_join(sample_size) %>%
  mutate(myaxis = paste0(species_range, " (n= ", num, ")")) %>%

  # plot
  ggplot(aes(x=myaxis, y=indicator1 , fill=species_range)) +
      geom_violin(width=1, linewidth = 0)  +
      geom_jitter(size=.5, width = 0.1) +
      xlab("") + ylab("Proportion of popuations with Ne > 500") +
      coord_flip() +
      scale_fill_manual(breaks=c("wide_ranging", "restricted", "unknown"),
                       labels=c("wide ranging", "restricted", "unknown"),
                       values=c("#F8766D", "#00BFC4", "grey80")) +
      theme_light() +
      theme(panel.border = element_blank(), legend.position="none", text= element_text(size=20))
```

## Indicator 1 by IUCN status

Indicator 1 by global IUCN:

```{r indicator1 gobalIUCN}

## Global IUCN
## prepare data
# add sampling size
sample_size <- indicators_full %>%
               filter(!is.na(indicator1)) %>% 
               group_by(global_IUCN) %>% summarize(num=n())

# new df 
df<- indicators_full %>% 
     filter(!is.na(indicator1)) %>% 
        # add sampling size 
        left_join(sample_size) %>%
        mutate(myaxis = paste0(global_IUCN, " (n= ", num, ")"))


# change order of levels so that they are in the desired order
df$myaxis<-factor(df$myaxis, 
                  #grep is used below to get the sample size, which may change depending on the data
                  levels=c(grep("cr", unique(df$myaxis), value = TRUE),
                          grep("en", unique(df$myaxis), value = TRUE),
                          grep("vu", unique(df$myaxis), value = TRUE),
                          grep("nt", unique(df$myaxis), value = TRUE),
                          grep("lc", unique(df$myaxis), value = TRUE),
                          grep("dd", unique(df$myaxis), value = TRUE),
                          grep("not_assessed", unique(df$myaxis), value = TRUE),
                          grep("unknown", unique(df$myaxis), value = TRUE)))

df$global_IUCN<-factor(df$global_IUCN, levels=c("cr", "en", "vu", "nt", "lc", "dd", "not_assessed", "unknown"))


      
# plot
df %>%
    ggplot(aes(x=myaxis, y=indicator1 , fill=global_IUCN)) +
      geom_violin(width=1.5, linewidth = 0)  +
      geom_jitter(size=.5, width = 0.1) +
      xlab("") + ylab("Proportion of popuations with Ne > 500") +
      coord_flip() +
      scale_fill_manual(values= IUCNcolors, # iucn color codes
                        breaks=c(levels(df$global_IUCN))) +
      theme_light() +
      ggtitle("global Red List") +
      theme(panel.border = element_blank(), legend.position="none", text= element_text(size=15))

```

Indicator 1 by regional IUCN (this may make no sense for your country if there is no IUCN redlist assesments at regional level:

```{r indicator1 regionalIUCN}
## Regional IUCN

## prepare data
# add sampling size
sample_size <- indicators_full %>%
               filter(!is.na(indicator1)) %>% 
               group_by(regional_redlist) %>% summarize(num=n())

# new df 
df<- indicators_full %>% 
     filter(!is.na(indicator1)) %>% 
        # add sampling size 
        left_join(sample_size) %>%
        mutate(myaxis = paste0(regional_redlist, " (n= ", num, ")"))

# change order of levels so that they are in the desired order
df$myaxis<-factor(df$myaxis, 
                  #grep is used below to get the sample size, which may change depending on the data
                  levels=c(grep("re", unique(df$myaxis), value = TRUE),
                          grep("cr", unique(df$myaxis), value = TRUE),
                          grep("en", unique(df$myaxis), value = TRUE),
                          grep("vu", unique(df$myaxis), value = TRUE),
                          grep("nt", unique(df$myaxis), value = TRUE),
                          grep("lc", unique(df$myaxis), value = TRUE),
                          grep("dd", unique(df$myaxis), value = TRUE),
                          grep("not_assessed", unique(df$myaxis), value = TRUE),
                          grep("unknown", unique(df$myaxis), value = TRUE)))

df$regional_redlist<-factor(df$regional_redlist, levels=c("re","cr", "en", "vu", "nt", "lc", "dd", "not_assessed", "unknown"))
      
# plot
df %>%
    ggplot(aes(x=myaxis, y=indicator1 , fill=regional_redlist)) +
      geom_violin(width=1, linewidth = 0)  +
      geom_jitter(size=.5, width = 0.1) +
      xlab("") + ylab("Proportion of popuations with Ne > 500") +
      coord_flip() +
      scale_fill_manual(values= IUCNcolors_regional, # iucn color codes
                        breaks=c(levels(df$regional_redlist))) +
      theme_light() +
      ggtitle("regional Red List") +
      theme(panel.border = element_blank(), legend.position="none", text= element_text(size=15))



```
